RL-Musician: A Configurable Open-Source Tool for Species Counterpoint Composition

### 2020-05-08 docs/paper/paper.tex
docs: Update first three sections of the draft

\date{April 14, 2020}

@@

\begin{abstract}
In this paper, automatic composition of first species counterpoint is framed as a tree search problem close to reinforcement learning. An environment representing a polyphonic piece in progress is introduced. An action is filling of the current measure (bar) for all melodic lines simultaneously. Once a specified number of measures is added, an episode terminates and the resulting piece is evaluated based on some hand-written rules derived from music theory. Best sequences of actions are found with a variation of Monte Carlo Beam Search.

@@

The word "species" in the name of the framework indicates that there are some types -- each of them has its own rhythmic pattern. Here, first species counterpoint is covered. This type has trivial rhythmic structure such that only whole (semibreve) notes can be used and every note must be placed in a beginning of a measure. Thus, interaction between rhythm and tonality is eliminated and the problem becomes even simpler.

Brief outline of the setup is as follows. There is an environment that keeps a piece in progress and the piece consists of a fixed number of measures. Each measure must be either empty (if it is not filled yet) or containing exactly one note per a melodic line. Initially, only the first measure and the last measure are filled with user-defined values and the current measure is the second one. An action is filling of the current measure, i.e., adding a pitch to each of the melodic lines. After an action, the next measure becomes the current one. When the last measure is reached, episode is ended and a reward based on evaluation of the created composition is assigned to the sequence of passed actions.

@@

These examples can be divided into two groups. The first one is formed by approaches where an explicitly given fitness function is optimized. The second group is formed by composition of music with reinforcement learning in its narrow sense (i.e., methods based on value function estimation). This group is something in between the first group and supervised learning approaches. On the one hand, reward function is a domain-specific name for fitness function. On the other hand, data generated during past episodes are used for training supervised models.

The closest to the current study works \cite{herremans2012first, herremans2013fifth} belong to the first group. A local search algorithm named Variable Neighborhood Search is used there in order to create first and fifth species counterpoint pieces that maximize fitness function. What these studies and the current one share in common is that fitness (reward) function is backed by the theory of species counterpoint. However, the current study is a bit closer to reinforcement learning, because composition of a piece is considered an episode consisting of sequential steps after each of which a list of valid actions changes. Conversely, in \cite{herremans2012first, herremans2013fifth}, the whole piece is created at random and then improved by applying small alterations. As a result, pieces that break some fundamental rules of counterpoint are also evaluated there (they are marked as unfeasible), whereas here such pieces cannot be created. This way of that, the core drawback of the two discussed papers is that download links to the software implementation named Optimuse are broken and so the results cannot be reproduced.

There are other optimization-based approaches as well. Instead of local search metaheuristics, population-based metaheuristics (genetic algorithms, evolutionary strategies, cross-entropy method), or construction metaheuristics (ant colony algorithms) can be used. However, such papers are less relevant to the current study and so let us switch our attention to reinforcement learning. It can be applied to algorithmic composition either as a standalone paradigm or in connection with supervised deep learning. As for the first case, probably, there are no influential works and only separate studies like \cite{yi2007automatic, smith2012reinforcement} are available. Conversely, the second group recently gained attention and several interesting approaches were developed there.

@@

The described in Section \ref{sec:introduction} procedure of first species counterpoint writing can be considered a tree search problem. The root is the initial state of an environment. Each valid option to fill the current measure is an edge from the current node of the tree to a new node. Leaves are pieces where all measures are filled.

If size of a tree is small enough, exhaustive search is possible. There are well-known classical algorithms such as depth-first search (DFS) that allow iterating over all paths from the root to leaves. Unfortunately, trees are large in numerous real problems. For instance, a rough estimate of total number of leaves for first species counterpoint with 2 voices and 16 measures is about $7^{14}$ which is equal to several hundreds of billions.

>>>>

\date{May 11, 2020}

@@

\begin{abstract}
In this paper, automatic composition of fifth species counterpoint is framed as a tree search problem close to reinforcement learning. The goal in this problem is to compose a melodic line given another melodic line consisting of whole notes only. The line in progress has fixed ends and the rest is filled sequentially from start to finish. Once it is done, an episode terminates and the resulting piece is evaluated based on some hand-written rules derived from music theory. Best sequences of notes are found with a variation of Monte Carlo Beam Search.

@@

The word "species" in the name of the framework indicates that there are some types -- each of them has its own rhythmic pattern. Here, fifth species counterpoint is covered. This type is the highest one and the less restrictive one. Like in other species, there is a given melodic line consisting entirely of whole (semibreve) notes. This line is called cantus firmus. Again, like in other species, a new melodic line of the same duration is going to be created (so called counterpoint line) and these two lines should sound aesthetically pleasant when played simultaneously. The difference between fifth species and other species is that a counterpoint line may change its rhythmic pattern: half notes, quarter notes, eights notes, and suspended whole notes can be used.

Brief outline of the setup is as follows. There is an environment that keeps a piece in progress. Initially, the piece consists of a user-defined cantus firmus, the first note of counterpoint line, and the last note of counterpoint line. An action is extension of the counterpoint left part by one new note. Not every action is valid, there are rules prohibiting some of them. Invalid actions can not be done by design. When the last note is reached, episode is ended and a reward based on evaluation of the created composition is assigned to the sequence of passed actions.

@@

These examples can be divided into two groups. The first one is formed by approaches where an explicitly given fitness function is optimized. The second group is formed by composition of music with reinforcement learning in its narrow sense (i.e., methods based on value function estimation). This group is something in between the first group and supervised learning approaches. On the one hand, reward function is a domain-specific name for fitness function. On the other hand, data generated during past episodes are used for training supervised models that approximate value function.

The closest to the current study works \cite{herremans2012first, herremans2013fifth} belong to the first group. A local search algorithm named Variable Neighborhood Search (VNS) is used there in order to create first and fifth species counterpoint pieces that maximize fitness function. What these studies and the current one share in common is that fitness (reward) function is backed by the theory of species counterpoint. Despite that, there are important differences:
\begin{itemize}
	\item In \cite{herremans2012first, herremans2013fifth}, the whole piece is created at random and then improved by applying small alterations. As a result, there are only evaluation rules and there are no rules of composition. If a piece breaks some fundamental rules of counterpoint, it is still evaluated (but it can receive score making it unfeasible). Conversely, in the current study, both rules of composition and rules of evaluation are present as separate entities. An action that contradicts a rule of composition can not be done and so only compliant pieces are evaluated.
	\item In \cite{herremans2012first, herremans2013fifth}, evaluation rules accurately follow counterpoint theory, but here it is not reproduced too literally. From a modern listener point of view, violation of counterpoint prescriptions does not necessarily sound unpleasant, because some of these prescriptions are style-specific.
	\item Output of VNS is a fixed piece. If cantus firmus is changed, VNS should be run from scratch. Actually, output of Monte Carlo Beam Search is a fixed piece too and search should be run from scratch for every new cantus firmus as well. However, the current approach can be upgraded to a true reinforcement learning approach. Such upgrade is possible, because the problem is formulated in RL terms: composition of a piece is considered an episode consisting of sequential steps after each of which a list of valid actions updates. If so, there can be an agent that observes current state of a piece and writes counterpoint line based on observations. Cantus firmus can be changed from episode to episode during training of the agent in order to teach the agent to deal with various cantus firmi. After each episode played by the trained agent, a new output piece can be created even for a fixed cantus firmus.
	\item Methodology of the two discussed papers has a software implementation named Optimuse, but download links are broken and so results can not be reproduced. Unlike this, links from the present paper lead to time-tested sites such as GitHub and PyPI.
\end{itemize}

There are other optimization-based approaches as well. Instead of local search metaheuristics, population-based metaheuristics (genetic algorithms, evolutionary strategies, cross-entropy method), or construction metaheuristics (ant colony algorithms) can be used. However, such papers are less relevant to the current study and so let us switch our attention to reinforcement learning. It can be applied to algorithmic composition either as a standalone paradigm or in connection with supervised deep learning. As for the first case, some separate studies like \cite{yi2007automatic, smith2012reinforcement} are available, but the second group recently gained attention and several interesting approaches were developed there.

@@

The described in Section \ref{sec:introduction} procedure of fifth species counterpoint writing can be considered a tree search problem. The root is the initial state of an environment. Each valid option to extend counterpoint line is an edge from the current node of the tree to a new node. Leaves are pieces where counterpoint line is finished.

If size of a tree is small enough, exhaustive search is possible. There are well-known classical algorithms such as depth-first search (DFS) that allow iterating over all paths from the root to leaves. Unfortunately, trees are large in numerous real problems.
